---
title: "README"
output: github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# gofedf

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/gofedf)](https://cran.r-project.org/package=gofedf) 
[![GitHub Release](https://img.shields.io/github/release/pnickchi/gofedf?style=flat)](https://github.com/pnickchi/gofedf/releases) 
[![Github All Releases](https://img.shields.io/github/downloads/pnickchi/gofedf/total.svg?style=flat)](https://github.com/pnickchi/gofedf)
<!-- badges: end -->


Authors:

*   [Richard Lockhart](http://www.sfu.ca/~lockhart/),
    <lockhart@sfu.ca>
*   [Payman Nickchi](https://github.com/pnickchi),
    <pnickchi@sfu.ca> (Maintainer)


The `gofedf` package provides tools to apply goodness-of-fit tests based on empirical distribution function theory. The package offers functions and routines to test the hypothesis that a univariate sample follows a distribution based on the empirical distribution function. The theory is founded on reducing the problem to a stochastic process and computing its covariance function. An approximate p-value, computed using the farebrother method based on the limiting distribution of the statistic, is provided. Users can run the test by calculating either the Cramer-von Mises or Anderson-Darling statistic. The covariance function of the stochastic process relies on specific characteristics of the assumed model. Notably, knowledge of the Fisher information matrix and the partial derivatives of the cumulative distribution function is crucial for computing the covariance function. However, obtaining these quantities can be computationally intensive or challenging in general likelihood models. To overcome this limitation, we propose an alternative method for estimating the covariance function of the stochastic process directly from the sample data. The package provides tools for this estimation and for testing if a sample comes from any general likelihood model. In summary, the package can be used to apply a goodness-of-fit test in any of the following settings:

1) Verify if the assumptions about the response variable in a generalized linear model (with any link function) are satisfied. The current version only checks for the Gamma distribution.

2) Validate if the residuals of a linear model follow a Normal distribution.

3) Apply a goodness-of-fit test to examine if a set of bivariate samples follows a Normal, Gamma, or Exponential distribution.

4) Determine if a bivariate continuous set of data conforms to a specific distribution defined by the user. In this case, probability inverted values of the sample and the score function (if any parameter estimation is involved) are required.

## Installation

You can install the released version of gofedf from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
install.packages('gofedf')
```


And the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github('pnickchi/gofedf')
```


## Examples


### 1. Bivariate Normal distribution
In this example, we demonstrate how to apply a goodness-of-fit test based on the empirical distribution function (EDF) to assess whether a vector of data follows a Normal distribution.

```{r, echo = FALSE}
library(gofedf)
```


```{r}
# Reproducible example
set.seed(123)

# Randomly generate some data
n <- 50
sim_data <- rnorm(n)

# Test if the data follows a Normal distribution, calculate Cramer-von Mises statistic and approximate p-value of the test.
testNormal(x = sim_data, method = 'cvm')

# Test if the data follows a Normal distribution, calculate Anderson-Darling statistic and approximate p-value of the test.
testNormal(x = sim_data, method = 'ad')

```


### 2. Bivariate Gamma distribution
In this example, we show how to apply goodness of fit test over a vector of data and check if the data follows a Gamma distribution.

```{r}
# Reproducible example
set.seed(123)

# Randomly generate some data
n <- 50
sim_data <- rgamma(n, shape = 3)

# Test if the data follows a Gamma distribution, calculate Cramer-von Mises statistic and approximate pvalue
testGamma(x = sim_data, method = 'cvm')
```


### 3. Linear model with normal residuals

```{r}
# Reproducible example
set.seed(123)

# Create a set of explanatory variables and response according to a linear model
n <- 50
p <- 5
X <- matrix( runif(n*p), nrow = n, ncol = p)
e <- runif(n)
b <- runif(p)
y <- X %*% b + e

# Test if the residuals of the model follows a Normal distribution, calculate Cramer-von Mises statistic and approximate pvalue
testLMNormal(x = X, y)

# Or alternatively just pass 'myfit' object directly instead of X and y:
# myfit <- lm(y ~ X - 1, x = TRUE, y = TRUE)
# testLMNormal(fit = myfit)
```


### 4. Generalized linear model with Gamma response and log link function

```{r}
# Reproducible example
set.seed(123)


# Create a set of explanatory variables and response according to a generalized linear model with log link
n <- 50
p <- 2
X <- matrix( rnorm(n*p, mean = 10, sd = 0.1), nrow = n, ncol = p)
b <- runif(p)
e <- rgamma(n, shape = 3)
y <- exp(X %*% b) * e

# Test if the residuals of the model follows a Gamma distribution, calculate Cramer-von Mises statistic and approximate pvalue
testGLMGamma(x=X, y, l = 'log', method = 'cvm')
```



### General likelihood model

One of the most important features of the package is to allow users to apply the goodness of fit test based on empirical distribution functions for any general likelihood model. This is helpful when the functions and procedures for a distribution is not included in any R packages. For example, imagine you have a sample of size $n$ such as $x_{1},x_{2},...,x_{n}$ from a model with $p$ unknown parameters such as $\theta$. The package requires two inputs to apply the test:

1) A function to calculate score values for each observation and returns a (n x p) matrix. Each rows calculates the score for each observation. Alternatively you might have the score values as a matrix which is also acceptable.     

2) A vector of length $n$ where elements where each element is $F^{-1}(x_{i})$.

```{r}
# Example: Inverse Gaussian (IG) distribution with weights

# Reproducible example
#set.seed(123)


# Set the sample size
n <- 50

# Assign weights
covariates <- rep(1.5,n)

# Set mean and shape parameters for IG distribution.
mio        <- 2
lambda     <- 2

# Generate a random sample from IG distribution with weighted shape.
y <- statmod::rinvgauss(n, mean = mio, shape = lambda * covariates)

# Compute MLE of parameters, score matrix, and pit values.
theta_hat    <- IG_mlefunc(obs = y, w = covariates)
score.matrix <- IG_scorefunc(obs = y, mle = theta_hat, w = covariates)
pit.values   <- IG_pitfunc(obs = y , mle = theta_hat)

# Apply the goodness-of-fit test.
testYourModel(x = y, pit = pit.values, score = score.matrix)


```


### References
[1] Stephens, M.A. (1974). [EDF Statistics for Goodness of Fit and Some Comparisons.] *Journal of the American Statistical Association*, Vol. 69, 730-737.
[2] Stephens, M.A. (1976). [Asymptotic results for goodness-of-fit statistics with unknown parameters.] *Annals of Statistics*, Vol. 4, 357-369.
