---
title: "README"
output: github_document
date: `Sys.Date()`
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# GoFTest

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/PEIMAN2)](https://cran.r-project.org/package=PEIMAN2) 
[![GitHub Release](https://img.shields.io/github/release/jafarilab/PEIMAN2?style=flat)](https://github.com/jafarilab/PEIMAN2/releases) [![Github All Releases](https://img.shields.io/github/downloads/jafarilab/PEIMAN2/total.svg?style=flat)](https://github.com/jafarilab/PEIMAN2)

<!-- badges: end -->


The GoFTest package provides tools to apply goodness of fit tests based on empirical distribution function theory. The software provides functions and routines to test the hypothesis that a sample follows a distribution by calculating Cramer-von Mises or Anderson-Darling statistic and computing the approximate pvalue by Imhof method.

The package can be used to test the following: 

1) Goodness of fit test to check if the residuals of a generalized linear model follows Gamma distribution.
2) Goodness of fit test to check if the residuals of a linear model follows Normal distribution.
3) Apply goodness of fit test for bivariate Normal, Gamma, Exponential.
4) Apply goodness of fit test for distributions that are not inclueded in R (distributions defined by


## Installation

You can install the released version of GoFTest from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
install.packages('GoFTest')
```


And the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github('pnickchi/GoFTest')
```


## Example

To be completed soon.
