---
title: "README"
output: github_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# GoFTest

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/PEIMAN2)](https://cran.r-project.org/package=PEIMAN2) 
[![GitHub Release](https://img.shields.io/github/release/jafarilab/PEIMAN2?style=flat)](https://github.com/jafarilab/PEIMAN2/releases) [![Github All Releases](https://img.shields.io/github/downloads/jafarilab/PEIMAN2/total.svg?style=flat)](https://github.com/jafarilab/PEIMAN2)

<!-- badges: end -->


Authors:

*   [Richard Lockhart](http://www.sfu.ca/~lockhart/),
    <lockhart@sfu.ca>
*   [Payman Nickchi](https://github.com/pnickchi),
    <pnickchi@sfu.ca> (Maintainer)


The GoFTest package provides tools to apply goodness of fit tests based on empirical distribution function theory. The software provides functions and routines to test the hypothesis that a sample follows a distribution by calculating Cramer-von Mises or Anderson-Darling statistic. An approximate p-value based on the limiting distribution of statistic
is computed by Imhof method. More importantly the package can be used to test if a sample comes from any general likelihood model. In summary, the package can be used to apply goodness of fit test in any of the following settings: 

1) Check if the assumptions about the response variable in a generalized linear model (with any link) is satisfied. Current
version only check for Gamma distribution.
2) Check if the residuals of a linear model follows a Normal distribution.
3) Apply goodness of fit test to examine if a set of bivariate sample follows Normal, Gamma, or Exponential distribution.
4) Check if a bivariate continuous set of data follows any specific distribution that is defined by user. In this case, probability inverted values of the sample and score function (if there is any parameter estimation involved) is needed.


## Installation

You can install the released version of GoFTest from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
install.packages('GoFTest')
```


And the development version from [GitHub](https://github.com/) with:

```{r, eval=FALSE}
# install.packages("devtools")
devtools::install_github('pnickchi/GoFTest')
```


## Example


### Bivariate Normal distribution
In this example, we show how to apply goodness of fit test over a vector of data and check if the data follows a normal distribution.

```{r, echo = FALSE}
library(GoFTest)
```


```{r}
# Reproducible example
set.seed(123)

# Randomly generate some data
n <- 50
sim_data <- rnorm(n)

# Test if the data follows a normal distribution, calculate Cramer-von Mises statistic and approximate pvalue
testNormal(x = sim_data, method = 'cvm')

# Test if the data follows a normal distribution, calculate Anderson-Darling statistic and approximate pvalue
testNormal(x = sim_data, method = 'ad')

```


### Bivariate Gamma distribution
In this example, we show how to apply goodness of fit test over a vector of data and check if the data follows a Gamma distribution.

```{r}
# Reproducible example
set.seed(123)

# Randomly generate some data
n <- 50
sim_data <- rgamma(n, shape = 3)

# Test if the data follows a Gamma distribution, calculate Cramer-von Mises statistic and approximate pvalue
testGamma(x = sim_data, method = 'cvm')
```


### Linear model with normal residuals

```{r}
# Reproducible example
set.seed(123)

# Create a set of explanatory variables and response according to a linear model
n <- 50
p <- 5
X <- matrix( runif(n*p), nrow = n, ncol = p)
e <- runif(n)
b <- runif(p)
y <- X %*% b + e

# Test if the residuals of the model follows a Normal distribution, calculate Cramer-von Mises statistic and approximate pvalue
testLMNormal(x = X, y)

# Or alternatively just pass 'myfit' object directly instead of X and y:
# myfit <- lm(y ~ X - 1, x = TRUE, y = TRUE)
# testLMNormal(fit = myfit)
```


### Generalized linear model with Gamma response and log link function

```{r}
# Reproducible example
set.seed(123)


# Create a set of explanatory variables and response according to a generalized linear model with log link
n <- 50
p <- 2
X <- matrix( rnorm(n*p, mean = 10, sd = 0.1), nrow = n, ncol = p)
b <- runif(p)
e <- rgamma(n, shape = 3)
y <- exp(X %*% b) * e

# Test if the residuals of the model follows a Gamma distribution, calculate Cramer-von Mises statistic and approximate pvalue
testGLMGamma(x=X, y, l = 'log', method = 'cvm')
```



### User defined distributions

One of the most important features of the package is to allow users to apply the goodness of fit test based on empirical distribution functions for ant general likelihood model. This is helpful when the functions and procedures for a distribution is not included in any R packages. For example, imagine you have a sample of size $n$ such as $x_{1},x_{2},...,x_{n}$ from a model with $p$ unknown parameters such as $\theta$. The package requires two inputs to apply the test:

1) A function to calculate score values for each observation and returns a (n x p) matrix. Each rows calculates the score for each observation. Alternatively you might have the score values as a matrix which is also acceptable.     

2) A vector of length $n$ where elements where each element is $F^{-1}(x_{i})$.

```{r}
# Example: Inverse Gaussian (IG) distribution with weights

# Reproducible example
#set.seed(123)


# Set the sample size
n <- 50

# Assign weights
covariates <- rep(1.5,n)

# Set mean and shape parameters for IG distribution.
mio        <- 2
lambda     <- 2

# Generate a random sample from IG distribution with weighted shape.
y <- statmod::rinvgauss(n, mean = mio, shape = lambda * covariates)

# Compute MLE of parameters, score matrix, and pit values.
theta_hat    <- IG_mlefunc(obs = y, w = covariates)
score.matrix <- IG_scorefunc(obs = y, mle = theta_hat, w = covariates)
pit.values   <- IG_pitfunc(obs = y , mle = theta_hat)

# Apply the goodness-of-fit test.
testYourModel(x = y, pit = pit.values, score = score.matrix)


```


